---
title: "Basic Parameter fitting Ram et al. Paper"
output:
  html_document:
    df_print: paged
---

```{r, include= FALSE}
#loading required packages and sourcing Grind
if(!require(pacman)) {
install.packages("pacman"); require(pacman)}
p_load(tidyverse, rootSolve, deSolve, FME)
source("grind.R")


```
#Question 2
Before we start to fit, let us have a look at the data

```{r}
# The 3 experiments are indexed by their date:
exptA <- "2015-11-18"  #Experiment A
exptB <- "2015-12-14"  #Experiment B
exptC <- "2016-01-06"  #Experiment C

#Let us read in data of Figure 3 for experiment A
fig3RA <- read.csv(paste("Fig3/",exptA,"_R.csv",sep="")) # Red strain
fig3GA <- read.csv(paste("Fig3/",exptA,"_G.csv",sep="")) # Green strain
plot(fig3RA$Time, fig3RA$OD, ylim=c(0,0.8),col="red",pch=".",xlab="Time (hr)",ylab="OD", main = "Figure 3 A1 (red) and A2 (green)")
points(fig3GA$Time, fig3GA$OD, ylim=c(0,0.8),col="green",pch=".")

fig4 <- read.csv(paste("Fig4/",exptA,"_RG.csv",sep=""))
plot(fig4$Time, fig4$OD, ylim=c(0,0.8), col="blue", pch=".", xlab="Time (hr)", ylab="Total OD", main = "Figure 4 A")

fig5 <- read.csv(paste("Fig5/flow_df_",exptA,".csv",sep=""))
fig5G <- fig5[fig5$Strain=="Green",]
fig5R <- fig5[fig5$Strain=="Red",]
plot(fig5G$time, fig5G$freq_mean, ylim=c(0,1), col="green",xlab="Time (hr)", ylab="Frequency", main =  "Figure 5 A")
points(fig5R$time, fig5R$freq_mean, col="red")


```


Now to fit we first need to define the model

```{r}

model <- function(t, state, parms) {
  with(as.list(c(state,parms)), {
    a <- q0/(q0+exp(-m*t))
    dN <- r*a*N*(1-(N/K)^v)
    return(list(dN))  
  }) 
}

s <- c(N=0.124)
p <- c(K=0.6,r=0.4,m=2,q0=0.005,v=2)
```


To fit, we need to do two more things:
1. Tell Grind which parameters we would like to fit to the data. All other parameters will be kept constant, at a previously defined value.
2. Tell Grind what the lower and/or upper bounds are for certain parameters. E.g. we don't want negative growth speeds, so r would need a lower bound of 0. Similarly, Ram et al., specifiy that v should be equal or larger than 1, so that would require a lower bound of 1. If nothing is specified the parameter fit can take on any value.

```{r}
#N + all the parameters are free to be fitted. 
free <- c("N",names(p))
#for the lower bound, we bound everything at 0, and v at 1.
#so you first use the length of the vector free to first set lower bound for all parameters to zero
#next for parameter v specifically you set the lower bound to 1
lower <- rep(0, length(free)); lower[match("v",free)] <- 1; lower  # set lower bounds


#to fit, Grind wants a data frame with time and the variables it needs to fit. Note how we rename OD to N to be consistent.
data3RA <- as.data.frame(cbind(fig3RA$Time,fig3RA$OD)); names(data3RA) <- c("time","N")

#now we fit the parameters. fun = log tells the fitting that we want to fit the log of the data, not the actual data. The other parameters are self-explanatory. Note that we also start from some initial guess for the parameter values and try to improve the fit from there. For now, we'll just use the parameters from above.
fit3RAlog  <- fit(data3RA, state = s, odes = model, parms = p,
               free=free, fun=log, lower=lower, pch=".",
               legend=FALSE, tstep=0.1, main="Fig 3 a red monoculture (fun=log)",
               ymin = 0, ymax = 1
               )
#in this fit, you can get at the fitted parameters using $par
print(fit3RAlog$par)

```

Now look at fitting statistics
```{r}
#Q2C&D

```

```{r}

#Q2D 
#here you can write code to do the same as above but now for the red strain in Fig 3B
#make sure to look at both parameter values and fit statistics


```


The below loads in all the data from Fig3 of the Ram paper(both their data and their fits). This data is in a list. You can look at sections of this list using $ or [[]]. This comes in handy when wanting to fit the model to multiple data sets at once. As for the plots of the standard fits, note that unfortunately Grind's standard time plot uses a specific colour palette, one for each ODE to plot, and since we fit green and red separately it is not easily possible to make the colours green for the green bacteria.

```{r}
source("../HelperFunctionsAnswerFile.R")
dataAndFitsFigureThree = readInDataFigureThreeRamPaper()

#you can access the data using either $ or [[]]:
head(dataAndFitsFigureThree$DataexptAFig3G)
head(dataAndFitsFigureThree[["DataexptAFig3R"]])

```
```{r}
#Question 3

#Q3A 
```

#Q3B
To do this, we need to tell the fit function:
-What parameters to fit for all data together (free parameters)
-What parameters to fit separately per dataset (different parameters).
-Which data sets to use for fitting

```{r}
#set up initial state, and initial guess for parameter values
s <- c(N=0.124)
p <- c(K=0.6,r=0.4,m=2,q0=0.005,v=2)
#all parameters are fitted
free <- c("N",names(p)) 
#define the parameters for which different values should be obtained from different data sets, for all other parameters a single fit value is obtained from the combined data sets
differ <- c("q0", "m")
#define the list of data sets to be fitted to
l <- list(dataAndFitsFigureThree$DataexptAFig3G, dataAndFitsFigureThree$DataexptAFig3R)
```

As before we need to set the lower bound of v to 1. But now, we are fitting 2*the parameters that differ per dataset, + all those that don't for all data together. So we need some code to make sure that we set the right parameter to have a lower bound of 1. To do that, we use the following:

```{r}
#the total number of parameters to fit = 2 * those in differ (as these need to be fitted twice, once
#for each data set) + all remaining parameters in free that should be fitted combined for the two datasets
totfree <- c(free[!(free %in% differ)], differ, differ)
npar <- length(totfree)
cat("Number of free parameters",npar)

#Now set all instances of v in the total number of parameters to have a lower bound of1, and all the other lower bounds to 0
lower <- rep(0,npar); lower[which(totfree == "v")] <- 1; lower  # set lower bounds
```

Now do the combined fitting

```{r}
fitq0mdiffer <- fit(data=l, free=free, differ=differ, fun=log, odes = model, state = s, parms = p, lower=lower, pch=".",legend=FALSE, tstep=0.1,  main="Red from A and B, q0 andd m differing", add=TRUE, ymin = 0, ymax = 1)
summary(fitq0mdiffer)
fitq0mdiffer$ssr

```

# Question 4

```{r}
#Q4A: define the 2D model here


```
Now let us read in the parameters as Ram et al. fitted them to figure 3A for the red and green strain
and let us construct a parameter vector for the 2D model from this

```{r}
# Take the parameters from the single fits (we now use the fitted values obtained in the paper). 
# Parameters are renamed, appending indices 1 and 2, respectively for the red (R) and green (G)
# strains.
pR        <- dataAndFitsFigureThree$FitexptAFig3R$par[2:length(dataAndFitsFigureThree$FitexptAFig3R$par)]
names(pR) <- paste(names(pR),"1",sep="")
pG        <- dataAndFitsFigureThree$FitexptAFig3G$par[2:length(dataAndFitsFigureThree$FitexptAFig3G$par)]
names(pG) <- paste(names(pG),"2",sep="")
#total parameter set to fit figure 4 is then pR+pG plus the competition parameters
p4        <- c(pR,pG,c1=1,c2=1)
```
Next construct initial conditions for N1 and N2 from the optical density data representing N1+N2

```{r}
# Determine the initial conditions from the data: optical density at start
# Assyne initially strains have equal frequency assigning half of OD to each.
fig4Data  <- readr::read_csv("Fig4/2015-11-18_RG.csv")
data4Fit  <- dplyr::bind_cols(fig4Data$Time, fig4Data$OD) %>% dplyr::rename(time = ...1, OD = ...2) %>% as.data.frame()
initialOD <- data4Fit[1,2]
s4        <- c(N1=initialOD/2,N2=initialOD/2);s4
```


```{r}
# Q4B Define which parameters are free to be fitted to the data of Fig4

```


```{r}
#Q4C Write code to do the actual fitting


```


```{r}
#Q4D redo the fitting for a 2D model with a single c parameter


```

