---
title: "Getting started with Ram model and Grind"
output: html_notebook
---

First we need to source the file grind.R which is a wrapper made around the deSolve, rootSolve etc packages 
```{r}
#loading required packages and sourcing Grind
if(!require(pacman)) {
install.packages("pacman"); require(pacman)}
p_load(tidyverse, rootSolve, deSolve, FME)
source("grind.R")
```

Below we define a function called model in a format appropriate for deSolve and Grind.
This function can be used to calculate the values of the differential equation dN at different time points through numerical integration. As arguments it uses t (for a vector of time points), state (for a vector containing the initial state of the variables in the system), and parms (for a vector containing the values of the parameters of the system). State and parameters should be named vectors.

The with(as.list(c(state, parms))) does nothing more than make these names available without having to write parms["v"] every time you want to use v from the parameters vector, and parms["q0"] every time you want to use q0.

```{r}
model <- function(t, state, parms) {
  with(as.list(c(state,parms)), {
    a <- q0/(q0+exp(-m*t))
    dN <- r*a*N*(1-(N/K)^v)
    return(list(dN))  
  }) 
}
```

below we specify the vector p with parameter values
```{r}
p <- c(K=0.6,r=0.4,m=2,q0=0.005,v=2)
```

next we specify the vector s with inital state of the variables
```{r}
s <- c(N = 0.124) #units are OD.
```

now we are going to numerically integrate the model, specifying the time interval, timestep of integration, initial state, parameter values and which model to solve
```{r}
run(tmax = 100, tstep = 1,
    state = s,
    parms = p,
    odes = model
    )
```


```{r}
