---
title: "Getting started with Ram model and Grind"
output: html_notebook
---

First we need to source the file grind.R which is a wrapper made around the deSolve, rootSolve etc packages 
```{r}
#loading required packages and sourcing Grind
if(!require(pacman)) {
install.packages("pacman"); require(pacman)}
p_load(tidyverse, rootSolve, deSolve, FME)
source("grind.R")
```

Below we define a function called model in a format appropriate for deSolve and Grind.
This function can be used to calculate the values of the differential equation dN at different time points through numerical integration. As arguments it uses t (for a vector of time points), state (for a vector containing the initial state of the variables in the system), and parms (for a vector containing the values of the parameters of the system). State and parameters should be named vectors.

The with(as.list(c(state, parms))) does nothing more than make these names available without having to write parms["v"] every time you want to use v from the parameters vector, and parms["q0"] every time you want to use q0.

```{r}
model <- function(t, state, parms) {
  with(as.list(c(state,parms)), {
    a <- q0/(q0+exp(-m*t))
    dN <- r*a*N*(1-(N/K)^v)
    return(list(dN))  
  }) 
}
```

below we specify the vector p with parameter values
```{r}
p <- c(K=0.6,r=0.4,m=2,q0=0.005,v=2)
```

next we specify the vector s with inital state of the variables
```{r}
s <- c(N = 0.124) #units are OD.
```

now we are going to numerically integrate the model, specifying the time interval, timestep of integration, initial state, parameter values and which model to solve
```{r}
run(tmax = 100, tstep = 1,
    state = s,
    parms = p,
    odes = model
    )
#Q1A&B answer:
#-if you change the parameters, r,K and v the delay function does not change, but population dynamics does: larger r also causes population to reach equilibrium faster, larger K results in higher equilibrium value, higher v causes a more rapid transition to equilibrium values.
```
Now we wanted to study this startup function:
Note that we do not simply write q0 and m, but access them through the parameter list
```{r}
#Q1C answer:
#function for q0/q0+e^(-m*t)
fDelay <- function(t, params = p) {params["q0"]/(params["q0"]+exp(-params["m"]*t))}

#make sure to give it a vector of times
times = seq(1, 100, 0.5)
output = fDelay(times)
dataFDelay = data.frame(time = times, output = output)
ggplot(dataFDelay, aes(x = time, y = output)) +
  geom_point(size = 2) +
  ggtitle("Delay equation values over time") +
  geom_line(size = 0.5)

#Q1D
# this function starts at q0/(q0+1) at time zero, which is small if q0 is small, and approaches 1 as time #progresses because of e^{-mt} going to zero. 
# the function thus ensures that effective growth rate is initially modulated by a factor smaller than 1, causing
# it to initially be lower; later on it is multiplied by 1 enabling maximum growth 
# if you change q0 from 0.005 to 0.5 or m from 2 to 20 this delay function approaches 1 earlier, for much smaller time values, this causes the population to reach its maximum growth speed and final stable size (much) faster.

```



```{r}
